---
title: "course_project"
author: "Bashir,I. (2022)"
date: "17/04/2022"
output: html_document
editor_options: 
  chunk_output_type: inline
---

# Linkage map construction

To start with the analysis, we load the data (here, we are using the recombinant inbred line **population Z003**) using the function `read.cross()` , and estimate the pairwise recombination fractions (and their respective LOD scores) using the function `est.rf()`:

```{r}
library(qtl)
maize <- read.cross(format="csv", file="data/population_Z003.csv", genotypes=c("0", "1", "2"), crosstype = "riself")
```

1: The exclusion of remaining heterozygotes in a RIL population (code==2 is the same as genotype=1); 2: The fact that the map is not estimated yet, so qtl package interprets each marker 10 cM apart (which is clearly wrong and we'll fix it).

```{r}
maize <- est.rf(cross = maize)
plotRF(maize, col.scheme = "redblue")
```

From the `plotRF()` function, we notice that 1106 markers have not been assigned to linkage groups yet. In practice, it is recommended to check marker missing data and segregation distortion before grouping, but we will skip it for the sake of time here.

### Marker grouping

Once we have the pairwise recombination fraction estimates, we can try to see which markers are in the same linkage group. In order to do so, we need to provide the maximum recombination fraction (`max.rf` argument) and minimum LOD score (`min.lod` argument). This values are provided to the function `formLinkageGroups()` and they are used to see whether two markers are linked or not while avoiding false positives. We will show where these values come from, but you can use them directly in your analysis.\
For `max.rf` , we can use something around 0.38, which is the value maximum recombination fraction of 0.50 when converted via Kosambi map function:

$$ 
m_ij = \frac{1}{4} log \binom {1+2r_ij}{1-2r_ij}
$$

A recombination fraction as big as 0.50 means that two markers are independently segregating (i.e. these two markers are not linked). See below:

```{r}
max.rf <- 0.38
kosambi <- function(r) (1/4)*log((1+(2*r))/(1-(2*r)))
kosambi(r =max.rf)
```

For `min.lod`, we can run Bonferroni correction on the number of tests that we have to perform in order to evaluate marker linkage. The number of tests is the number of marker pairs that we have in our data. As a first guess, we have:

```{r}
(M <- totmar(maize)) # number of markers
```

```{r}
(num.pair <- choose(M,2)) # number of marker pairs
```

```{r}
(min.lrt <- qchisq(1-(0.05/num.pair), 1)) # min LRT to consider two markers
```

```{r}
(min.lod <- 0.2174 * min.lrt) # conversion of LRT to LOD
```

Now, it is time to see how many linkage groups we have. First, we run the function `formLinkageGroups()` with `reorgMarkers = FALSE` just to see how the markers get distributed along the formed linkage groups:

```{r}
maize <- formLinkageGroups(maize, max.rf = 0.38, min.lod = 6.25, reorgMarkers = TRUE)
table(lg[,2])
```

We note that there are 10 linkage groups (as expected for maize). Now, we can use `reorgMarkers = TRUE` and update our cross object named maize with each marker numbered according to the linkage group it belongs

```{r}
plotRF(maize, col.scheme = "redblue")
```

The heatmap shows grouped but not yet ordered markers within each linkage group

## Marker ordering

We are going to show two ways of ordering markers. The first way uses the `orderMarkers()` function by `R/qtl` and usually needs some manual curation. The second way uses the MDS algorithm and it is faster and usually more accurate. You can choose which method you want to use and skip to its specific section, meaning that you do not need to run both ways.

**Using `orderMarkers` function by `R/qtl`:**\
`R/qtl` has a function that runs the Branch-and-Bound algorithm as an possible solution for the Traveling\
Salesman Problem (TSP) that ordering markers is. It usually provides a good solution. The problem is that\
Branch-and-Bound is very sensitive to the marker choice that is made to start the algorithm. Therefore, we run it at least a couple of times such that some effect of the first choices can be evaluated.\
We save the maize object under two new object called `maize.bb1` and `maize.bb2` , so we can update the\
with the results of two runs of the Branch-and-Bound algorithm. In addition, we initialize two objects that will store the log-likelihood of the ordering for each linkage group from both runs:

```{r}
maize.bb1 <- maize
maize.bb2 <- maize
loglik.bb1 <- loglik.bb2 <- c()
```

**Linkage groups**

The object c stores the number of the chromosome under evaluation.

The argument `chr` let us take a closer look into the `heatmap` of a specific chromosome, whose markers are clearly not ordered

```{r}
for(i in 1:10) {
  c <- i
  plotRF(maize, chr=c, col.scheme = "redblue")
  maize.bb1 <- orderMarkers(cross = maize.bb1, chr=c, use.ripple = FALSE, map.function="kosambi")
  print(pull.map(maize.bb1, chr = c))
  plotRF(maize.bb1, chr = c, col.scheme = "redblue")
  print(loglik.bb1[c] <- attr(maize.bb1$geno[[c]]$map, "loglik"))
  maize.bb2 <- orderMarkers(cross = maize.bb2, chr=c, use.ripple = FALSE, map.function="kosambi")
  print(pull.map(maize.bb2, chr = c))
  print(loglik.bb2[c] <- attr(maize.bb2$geno[[c]]$map, "loglik"))
}
```

### **Manual adjustments from `orderMarkers()` output**

Before any manual adjustments to the marker orderings, we need to check which runs of `orderMarkers()`performed best for each linkage group. In order to do so, we should look at linkage group lengths, maximum space between two consecutive markers (a.k.a. gap) and their log-likelihoods. Higher log-likelihoods indicate better (i.e. more likely) maps.\
We combine the information from the function `summaryMap()` with the one gathered in the objects `loglik.bb1` and `loglik.bb1` from each respective run:

```{r}
knitr::kable(cbind(summaryMap(maize.bb1), log.likelihood=c(loglik.bb1, sum(loglik.bb1))))
```

```{r}
plotMap(maize.bb1)
```

```{r}
plotRF(maize.bb1, col.scheme = "redblue")
```

```{r}
knitr::kable(cbind(summaryMap(maize.bb2), log.likelihood=c(loglik.bb2, sum(loglik.bb2))))
```

```{r}
plotMap(maize.bb2)
```

```{r}
plotRF(maize.bb2, col.scheme = "redblue")
```

```{r}
save.image("maize_z003.RData")
```

Now, we can select the best order so far, that we can still try to improve by making manual adjustments for each linkage group. As an strategy to improve marker ordering, we will find where the major gaps are, and fix it by moving the block of markers to its most likely position when looking at the `heatmap`.

**Linkage group 1**

From the map summary tables, we notice that from the first run of Branch-and-Bound (`maize.bb1`) provided a better map than the second run (-2697\>-2711 ) even though its length was slightly greater (215\>214 ).

```{r}
plotRF(maize.bb1, chr=1, col.scheme = "redblue")
```

By looking at its `heatmap`, there seems to be small inversions, but no bigger inversions that can be easily corrected manually, so we keep it as is. Small inversions can be fixed by the function `ripple()`. However, this function is not well optimized in R/qtl for many markers, so we can avoid it at this point.\
**Linkage group 2**\
For linkage group 2, the first run also provided greater log-likelihood (-2006\>-2016), with a length of 259cM and a gap as big as 91 cM, so we need to fix it. From the heatmap, we notice that there are three subgroups of markers, where the one in the middle seems misplaced. We locate the markers on each side by looking at the major gaps.

```{r}
plotRF(maize.bb1, chr = 2, col.scheme = "redblue")
```

```{r}
(gaps <- tail(sort(diff(maize.bb1$geno[[2]]$map)), 2))
```

```{r}
match(names(gaps), names(maize.bb1$geno[[2]]$map))
```

```{r}
maize.bb1 <- switch.order(maize.bb1, chr = 2, order = c(1:10,35:139,34:11), error.prob = 0)
```

```{r}
pull.map(maize.bb1, chr = 2)
```

```{r}
plotRF(maize.bb1, chr = 2, col.scheme = "redblue")
```

```{r}
(loglik.bb1[2] <- attr(maize.bb1$geno[[2]]$map, "loglik"))
```

And we can keep doing this for every linkage group. However, another alternative delivers better ordering in general much more efficiently.

## Multidimensional Scale (MDS) by MDSMap + MAPpoly

MDSMap implements the MDS algorithm, and MAPpoly has a nice wrapper function called `mds_mappoly`. This function only needs the recombination fraction matrix estimated by R/qtl, and it returns a new order that can be used by `switch.order()` R/qtl function to yield a newly ordered linkage group.\
First, we need to install MAPpoly package and use the function below to get the order given by the MDS algorithm for a chosen linkage group. You need to run it first so it becomes available into R before you actually use it:

```{r}
library(mappoly)
getMDSorder <- function(cross, chr){
  markers <- match(names(cross$geno[[chr]]$map), colnames(cross$rf))
  mat <- cross$rf[markers,markers]
  rec.mat <- lod.mat <- matrix(rep(NA, length(markers)^2), nrow =length(markers))
  colnames(rec.mat)<-colnames(lod.mat)<-rownames(rec.mat)<-rownames(lod.mat)<-colnames(mat)
  lod.mat[upper.tri(lod.mat)] <- mat[upper.tri(mat)]
  lod.mat[lower.tri(lod.mat)] <- t(lod.mat)[lower.tri(lod.mat)]#; image(lod.mat)
  rec.mat[lower.tri(rec.mat)] <- mat[lower.tri(mat)]
  rec.mat[upper.tri(rec.mat)] <- t(rec.mat)[upper.tri(rec.mat)]#; image(rec.mat)
  input.mat <- NULL
  input.mat$rec.mat <- rec.mat
  input.mat$lod.mat <- lod.mat
  mds.map <- mappoly::mds_mappoly(input.mat)
  mds.ord <- match(as.character(mds.map$locimap$locus), colnames(mat))
  return(mds.ord)
}
```

We'll create a new object called `maize.mds` which is a copy of our original cross object maize , so that we can update the ordering within `maize.mds` only. In addition, we'll create an empty object called `loglik.mds` to store the log-likelihood of the orderings obtained using MDS:

```{r}
maize.mds <- maize
loglik.mds <- c()
```

### Linkage groups 1-10

```{r}
for(i in 1:10) {
  c <- i
  mds.ord <- getMDSorder(cross = maize.mds, chr = c)
  maize.mds <- switch.order(cross = maize.mds, chr=c, order=mds.ord, maxit = 10000, tol=1e-5)
  plotRF(maize.mds, chr=c, col.scheme = "redblue")
  pull.map(maize.mds, chr = c)
  (loglik.mds[c] <- attr(maize.mds$geno[[c]]$map, "loglik"))}
```

```{r}
pull.map(maize.mds, chr = c)
```

```{r}
loglik.mds[c] <- attr(maize.mds$geno[[c]]$map, "loglik")
```

```{r}
knitr::kable(cbind(summaryMap(maize.mds), log.likelihood=c(loglik.mds, sum(loglik.mds))))
```

```{r}
plotMap(maize.mds)
```

```{r}
plotRF(maize.mds, col.scheme = "redblue")
```

\

\

\

\

\

\
