---
title: "Project RIL population_Z0??"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, tidy = TRUE)
```

Packages to be installed (only needed to be run once):

```{r, eval = FALSE}
# install.packages(c("qtl", "mappoly"))
```

# Introduction

[Add an introduction. For example, you can explain the importance of the crop, the origin of the population, the meaning of plant height for maize, and how the different QTL mapping approaches might help us to understand better the genetic basis of a trait and guide breeding decisions.]

# Material and Methods + Results

## Linkage map

Loading data:

```{r, warning = FALSE}
library(qtl)
maize <- read.cross(format = "csv", file = "population_Z0??.csv", 
                    genotypes = c("0", "1", "2"), crosstype = "riself")
```

Data summary:

```{r, warning = FALSE}
summary(maize)
geno.image(maize)
```

Chi-squared tests for segregation distortion:

```{r}
gt.maize <- geno.table(maize) 
head(gt.maize)
table(gt.maize$P.value < 0.05)
table(gt.maize$P.value < 0.05/totmar(maize))
```

Two-point recombination fraction calculation:

```{r}
maize <- est.rf(maize)
dim(maize$rf)
plotRF(maize, col.scheme = "redblue")
```

Grouping:

```{r, warning = FALSE}
maize <- formLinkageGroups(maize, max.rf = 0.38, min.lod = 6.25, reorgMarkers = TRUE)
summary(maize)
plotRF(maize, col.scheme = "redblue")
```

Ordering:

```{r}
getMDSorder <- function(cross, chr) {
    markers <- match(names(cross$geno[[chr]]$map), colnames(cross$rf))
    mat <- cross$rf[markers, markers]
    rec.mat <- lod.mat <- matrix(rep(NA, length(markers)^2), nrow = length(markers))
    colnames(rec.mat) <- colnames(lod.mat) <- rownames(rec.mat) <- rownames(lod.mat) <- colnames(mat)
    lod.mat[upper.tri(lod.mat)] <- mat[upper.tri(mat)]
    lod.mat[lower.tri(lod.mat)] <- t(lod.mat)[lower.tri(lod.mat)]  #; image(lod.mat)
    rec.mat[lower.tri(rec.mat)] <- mat[lower.tri(mat)]
    rec.mat[upper.tri(rec.mat)] <- t(rec.mat)[upper.tri(rec.mat)]  #; image(rec.mat)
    input.mat <- NULL
    input.mat$rec.mat <- rec.mat
    input.mat$lod.mat <- lod.mat
    mds.map <- mappoly::mds_mappoly(input.mat)
    mds.ord <- match(as.character(mds.map$locimap$locus), colnames(mat))
    return(mds.ord)
}
maize.mds <- maize
```

```{r, warning = FALSE}
for(c in 1:10) { # Reordering markers into each of the 10 linkage groups
    mds.ord <- getMDSorder(cross = maize.mds, chr = c)
    maize.mds <- switch.order(cross = maize.mds, chr = c, order = mds.ord)
    plotRF(maize.mds, chr = c, col.scheme = "redblue")
}
save(maize.mds, file = "maize_mds.RData")
```

Final map:

```{r}
summaryMap(maize.mds)
plotMap(maize.mds)
plotRF(maize.mds, col.scheme = "redblue")
```

## QTL mapping

Approximate critical values previously obtained from permutation tests (ideally, should be run separately for each population):

```{r}
lod.mr <- 3.0
lod.im <- 3.0
lod.cim <- 7.4
penalties.mim <- c(main = 2.92, heavy = 4.21, light = 2.05)
```

### Single marker analysis (SMA)

```{r}
maize.mr <- scanone(maize.mds, pheno.col = "PlantHeight", method = "mr")
summary(maize.mr, threshold = lod.mr)
plot(maize.mr, type = "p", main = "Single marker analysis (SMA)")
abline(h = lod.mr)
```

### Interval mapping (IM)

```{r}
maize.mds <- calc.genoprob(cross = maize.mds, step = 1)
```

```{r}
maize.im <- scanone(cross = maize.mds, pheno.col = "PlantHeight", method = "hk")
summary(maize.im, threshold = lod.im)
plot(maize.im, col = "blue", main = "Interval mapping (IM)")
abline(h = lod.im)
```

### Composite interval mapping (CIM)

```{r, warning = FALSE}
maize.cim10 <- cim(cross = maize.mds, pheno.col = "PlantHeight", 
                   method = "hk", n.marcovar = 20, window = 10)
(maize.cim.sig <- summary(maize.cim10, threshold = lod.cim))
plot(maize.cim10, col = "red", main = "Composite interval mapping (CIM)")
abline(h = lod.cim)
```

```{r}
maize.mds <- sim.geno(maize.mds, step = 1)
maize.cim.qtl <- makeqtl(cross = maize.mds, chr = maize.cim.sig$chr, 
                         pos = maize.cim.sig$pos)
maize.cim.qtl
```

```{r}
formula <- as.formula(paste("y", paste0("Q", c(1:length(maize.cim.sig$pos)), 
                                        collapse = "+"), sep = " ~ "))
maize.cim.fit <- fitqtl(cross = maize.mds, pheno.col = "PlantHeight", 
                        qtl = maize.cim.qtl, formula = formula, get.ests = TRUE)
summary(maize.cim.fit)
```

### Multiple interval mapping (MIM)

```{r}
maize.step <- stepwiseqtl(maize.mds, pheno.col = "PlantHeight", 
                          max.qtl = 6, method = "hk", penalties = penalties.mim)
summary(maize.step)
plotLodProfile(maize.step, main = "PlantHeight")
```

```{r}
maize.mds <- sim.geno(maize.mds, step = 5)
maize.two <- scantwo(maize.mds, pheno.col = "PlantHeight", method = "hk")
save(maize.two, file = "maize_two.RData")
```

```{r}
plot(maize.two, col.scheme = "redblue")
summary(maize.two)
```

# Discussion

[Add a discussion.]

# Conclusion

[Add a conclusion.]

# References

[Add references.]