---
title: "w12_OneMap"
author: "Bashir,I. (2022)"
date: "26/05/2022"
output: html_document
editor_options: 
  chunk_output_type: inline
---

## Genetic analyses of full-sib progenies using OneMap and fullsibQTL

### Linkage map construction

To build the linkage map, we are going to use the `OneMap` package, but we data that exists within the `fullsibQTL` package, so we need to install and load both of them.

```{r}
library(onemap)
```

```{r}
library(devtools)
devtools::install_github("guilherme-pereira/fullsibQTL")
```

```{r}
#install_github("augusto-garcia/fullsibQTL")
detach("package:fullsibQTL", unload=TRUE)
library(fullsibQTL)
```

```{r}
inputfile <- system.file("extdata", "example_QTLfullsib.raw", package = "fullsibQTL")
fs_data <- read_onemap(inputfile = inputfile)
```

There are some functions associated with such data that help us to visualize and understand it better:

```{r}
print(fs_data)
```

```{r}
plot(fs_data)
plot(fs_data, all = FALSE)
plot_by_segreg_type(fs_data)
```

Our fist step of the linkage analysis starts with segregation tests. We use the function `test_segregation()` to compute the chi-square as its associated value for each marker, depending on their expected segregation:

```{r}
segreg_test <- test_segregation(fs_data)
head(print(segreg_test))
```

```{r}
plot(segreg_test)
```

```{r}
no_dist <- select_segreg(segreg_test, distorted = FALSE, numbers = TRUE)
```

Most markers do follow the expected segregation, but 18% of the markers is distorted. We decide to proceed with the analyses without them.\
Next, we estimate the pairwise recombination fraction (two-point estimation):

```{r}
twopts <- rf_2pts(fs_data)
twopts
```

```{r}
print(twopts, c("M1", "M2"))
```

```{r}
mark_no_dist <- make_seq(twopts, c(no_dist))
```

And use the sequence of markers created by the function `make_seq()` to group them into linkage groups by using the function `group()`

```{r}
LOD_sug <- suggest_lod(fs_data)
LOD_sug
```

```{r}
LGs <- group(mark_no_dist, LOD = LOD_sug, max.rf = 0.37)
```

```{r}
print(LGs, detailed = FALSE)
```

We define the Kosambi map function using the `set_map_fun()` , and estimate the final multipoint genetic map (recombination fractions computed via hidden Markov models):

```{r}
set_map_fun(type = "kosambi")
LG1 <- make_seq(LGs, 1)
LG2 <- make_seq(LGs, 2)
LG3 <- make_seq(LGs, 3)
LG4 <- make_seq(LGs, 4)
LG1 <- map(LG1)
LG2 <- map(LG2)
LG3 <- map(LG3)
LG4 <- map(LG4)
```

Typically, we would need to go through a set of functions to order the markers within each linkage group. However, for the sake of time, we use the given order, which seems to be the correct one:

```{r}
rf_graph_table(LG1)
```

```{r}
rf_graph_table(LG2)
```

```{r}
rf_graph_table(LG3)
```

```{r}
rf_graph_table(LG4)
```

Finally, we can plot the map using the function draw_map() :

### QTL mapping

To perform QTL mapping, we are going to use the `fullsibQTL` package.\
First, we create an object that combines the raw data fs_data use by `OneMap` and the map information:

```{r}
fs_map <- create_fullsib(fs_data, map.list = list(LG1, LG2, LG3, LG4), step = 1,map.function = "kosambi")
fs_map
```

### Interval mapping

As always, we can initially try to find QTL using the interval mapping model below:\
This model tests if any of the possible effects of a single QTL is significant:

```{r}
im <- im_scan(fullsib = fs_map, pheno.col = 1)
summary(im)
```

To declare QTL, we perform the 1,000 permutation tests:

```{r}
set.seed(1234)
im_perm <- im_scan(fs_map, pheno.col = 1, n.perm = 100)
save(im_perm, file = "im_perm.Rdata")
```

Then, use the $a$ = 0.05 to be our threshold:

```{r}
load("im_perm.Rdata")
summary(im_perm, alpha = 0.05)
```

```{r}
im_thre <- summary(im_perm, alpha = 0.05, verbose = FALSE)[1, 1]
plot(im)
abline(h = im_thre, col = "red")
```

### Composite interval mapping

If we want to increase statistical power, we can use the composite interval mapping model below:\
The cofactors (marker covariates) - 10 in our case - are selected using the `cof_selection()` function

```{r}
cofs_fs <- cof_selection(fs_map, pheno.col = 1, k = log(300), n.cofactor = 10)
```

```{r}
plot(cofs_fs)
```

Then, we can proceed with the CIM scan:

```{r}
cim <- cim_scan(fullsib = cofs_fs, ws = 10, pheno.col = 1)
```

Again, 1,000 permutation tests is used to define a threshold:\

```{r}
cim_perm <- cim_scan(fullsib = cofs_fs, pheno.col = 1, n.perm = 1000)
save(cim_perm, file = "cim_perm.Rdata")
```

```{r}
load("cim_perm.Rdata")
summary(cim_perm, alpha = 0.05) 
```

```{r}
cim_thre <- summary(cim_perm, alpha = 0.05, verbose = FALSE)[1, 1]
plot(cim, lwd=1)
plot(im, lty=2, lwd=1 add=T)
abline(h = cim_thre, col = "red")
abline(h = im_thre, col = "red", lty=2)
```

Now we can compare the IM and CIM

```{r}
summary(cim)
```

The function `summary()` only shows the highest peak per chromosome. The function `plot_fullsibQTL()` can help us to identify the QTL peaks interactively

```{r}
plot_fullsibQTL(fullsib = fs_map, fullsib.scan = cim, interact = TRUE)
```

Given the information from the plot above (not shown), we end up with eight QTL:

```{r}
QTL1 <- cim_char(fullsib = cofs_fs, pheno.col = 1, lg = 1, pos = 15)
QTL2 <- cim_char(fullsib = cofs_fs, pheno.col = 1, lg = 1, pos = 125)
QTL3 <- cim_char(fullsib = cofs_fs, pheno.col = 1, lg = 2, pos = 21)
QTL4 <- cim_char(fullsib = cofs_fs, pheno.col = 1, lg = 2, pos = 61)
QTL5 <- cim_char(fullsib = cofs_fs, pheno.col = 1, lg = 2, pos = 87)
QTL6 <- cim_char(fullsib = cofs_fs, pheno.col = 1, lg = 3, pos = 23)
QTL7 <- cim_char(fullsib = cofs_fs, pheno.col = 1, lg = 3, pos = 71)
QTL8 <- cim_char(fullsib = cofs_fs, pheno.col = 1, lg = 4, pos = 45)
```

```{r}
knitr::kable(cbind(QTL1, QTL2, QTL3, QTL4, QTL5, QTL6, QTL7, QTL8))
```

Segregation and phase of each QTL can be obtained from the functions get_segr() and draw_phase() , respectively:

```{r}
get_segr(QTL1)
```

```{r}
draw_phase(fullsib = cofs_fs, fullsib.char = QTL1)
```

```{r}
get_segr(QTL2)
```

```{r}
draw_phase(fullsib = cofs_fs, fullsib.char = QTL2)
```

```{r}
get_segr(QTL3)
```

```{r}
draw_phase(fullsib = cofs_fs, fullsib.char = QTL3)
```

```{r}
draw_phase(fullsib = cofs_fs, fullsib.char = QTL4)
```

```{r}
get_segr(QTL5)
```

```{r}
draw_phase(fullsib = cofs_fs, fullsib.char = QTL5)
```

```{r}
get_segr(QTL6)
```

```{r}
draw_phase(fullsib = cofs_fs, fullsib.char = QTL6)
```

```{r}
get_segr(QTL7)
```

```{r}
draw_phase(fullsib = cofs_fs, fullsib.char = QTL7)
```

```{r}
get_segr(QTL8)
```

```{r}
draw_phase(fullsib = cofs_fs, fullsib.char = QTL8)
```

The $R^2$ for each QTL and all of them combined is computed by the function `r2_ls()` below:

```{r}
qtls.cim <- r2_ls(fs_map, pheno.col = 1, lg = c(1, 1, 2, 2, 2, 3, 3, 4), pos = c("M3",
"M14", "M21", "M25", "M27", "M33", "M38", "M53"))
knitr::kable(qtls.cim)
```

\

\

\

\
\

\

\

\

\

\

\

\

\

\

\

\
