---
title: "Week_9 Assignment"
author: "Bashir,I. (2022)"
date: "12/05/2022"
output: html_document
---

#### **Assignment - [QTL mapping III: composite interval mapping (CIM)](https://ava.ufv.br/mod/resource/view.php?id=146487 "QTL mapping III: composite interval mapping (CIM)")**

Use the linkage map to perform composite interval mapping (CIM) for the trait "PlantHeight". Do not forget to run permutation tests to identify the LOD threshold to declare QTL at an alpha level of 0.05. Results for the QTL peak within each QTL region should be presented as a table showing QTL chromosome and location of its peak, their effects, associated LOD scores, and R-squared.

CIM

For running CIM, there is an function called \`cim()\` . In this function, we need to specify two main arguments:

\- The number of cofactors or covariate markers: \`n.marcovar\` , which can be given as .

\- The window size: window , which can be different values depending on population size and map\\

saturation. It generally goes from 5 to 15 cM, but it can go as high as 20 or 30 cM, or take the whole\\

chromosome (i.e. leave one chromosome out - LOCO)

```{r}
maize.cim10 <- cim(cross = maize.mds, pheno.col = "PlantHeight", method = "hk", n.marcovar = 2 * sqrt(nind(maize.mds)), window = 10)
maize.cim15 <- cim(cross = maize.mds, pheno.col = "PlantHeight", method = "hk", n.marcovar = 2 * sqrt(nind(maize.mds)), window = 15)
maize.cim20 <- cim(cross = maize.mds, pheno.col = "PlantHeight", method = "hk", n.marcovar = 2 * sqrt(nind(maize.mds)), window = 20)
maize.cimInf <- cim(cross = maize.mds, pheno.col = "PlantHeight", method = "hk", n.marcovar = 2 * sqrt(nind(maize.mds)), window = Inf)
```

```{r}
summary(maize.cim10)
summary(maize.cim15)
summary(maize.cim20)
summary(maize.cimInf)
```

In order to make permutation tests to be less conservative and make them work for any window size, we can choose window \`= Inf\` . That means that no cofactors will be allowed withing the linkage group where the tests are being performed:

```{r}
maize.perm.cim <- cim(cross = maize.mds, pheno.col = "PlantHeight", method = "hk", n.marcovar = 2 * sqrt(nind(maize.mds)), window = Inf, n.perm = 1000)
```

```{r}
summary(maize.perm.cim, alpha = 0.05)
```

Now, if we use that permutation results to identify the most significant QTL, we have:

```{r}
summary(maize.cim10, perms = maize.perm.cim, alpha = 0.05)
summary(maize.cim15, perms = maize.perm.cim, alpha = 0.05)
summary(maize.cim20, perms = maize.perm.cim, alpha = 0.05)
summary(maize.cimInf, perms = maize.perm.cim, alpha = 0.05)
```

```{r}
plot(maize.cim10, maize.cim15, maize.cimInf, col = c("blue", "orange", "cyan"), main = "Composite interval mapping (CIM)")
add.threshold(maize.cim, perms = maize.perm.cim, alpha = 0.05)
legend("topright", legend = c("ws = 10", "ws = 15", "ws = Inf"), lty = 1, col = c("blue", "orange", "cyan"))
```

Because \`maize.cim20\` failed to declare most of QTL we think are true, and \`maize.cim15\` has the same results as \`maize.cimInf\` , we dropped them from the plot. We can compare our CIM results when window = 10 and window = Inf with our IM results:

```{r}
plot(maize.im, maize.cim10, maize.cimInf, col = c("red", "blue", "cyan"), main = "IM versus CIM")
add.threshold(maize.cim, perms = maize.perm.cim, alpha = 0.05)
add.threshold(maize.im, perms = maize.perm.im, alpha = 0.05, col = "red")
legend("topright", legend = c("IM", "CIM (ws = 10)", "CIM (ws = Inf)"), lty = 1, col = c("red", "blue", "cyan"))
```

We can also maybe limit the visualization to chromosomes where QTL appeared in the IM and CIM analyses:

```{r}
plot(maize.im, maize.cim10, maize.cimInf, col = c("red", "blue", "cyan"), chr = c(1, 2, 3, 7), main = "IM versus CIM")
add.threshold(maize.im, perms = maize.perm.im, alpha = 0.05, col = "red")
add.threshold(maize.cim10, perms = maize.perm.cim, alpha = 0.05, col = "black")
add.cim.covar(maize.cimInf, chr = c(1, 2, 3, 7), col = "green")
legend("topleft", legend = c("IM", "CIM (ws = 10)", "CIM (ws = Inf)"), lty = 1, col = c("red","blue", "cyan"))
```

The green dots represent the location of markers selected as covariates (cofactors) for the CIM search with window = Inf .\\

Noticed that the QTL on chromosome 1 and 3 that we had identified using IM is right below the threshold for the CIM analysis - so we called them suggetive QTL here. Depending on the researcher's decision, we could low the threshold a bit to include it as a QTL using CIM, which is probably okay. We'll see, however, a better way to deal with it when running a multiple-QTL model, which is our next topic.

Right now, suppose we want to investigate the QTL estimates under CIM with window = 10 . We just need to store the QTL chromosomes and positions from our selected model (CIM with )

```{r}
maize.cim.sig <- summary(maize.cim10, perms = maize.perm.cim, alpha = 0.05)
maize.cim.sig
```

However, we noticed that only one QTL on chromosome 1 is listed (the one with the highest LOD score). We need to find the other one by looking at the positions in that chromosome that have a LOD score greater than our threshold of 7.39:

```{r}
which(maize.cim10$lod[maize.cim10$chr == 1] > 7.39)
```

```{r}
maize.cim10$lod[which(maize.cim10$lod[maize.cim10$chr == 1] > 7.39)]
```

```{r}
maize.cim10$pos[which(maize.cim10$lod[maize.cim10$chr == 1] > 7.39)]
```

```{r}
chr <- c(1, maize.cim.sig$chr)
pos <- c(198, maize.cim.sig$pos)
```

In order to show the QTL effect estimates for an RIL population, we need to use the function `sim.geno()`, then `makeqtl()` and `fitqtl()`:

```{r}
maize.mds <- sim.geno(maize.mds, step = 1)
maize.cim.qtl <- makeqtl(cross = maize.mds, chr = chr, pos = pos)
maize.cim.qtl
```

```{r}
maize.cim.fit <- fitqtl(cross = maize.mds, pheno.col = "PlantHeight", qtl = maize.cim.qtl,formula = y ~ Q1 + Q2 + Q3 + Q4, get.ests = TRUE)
summary(maize.cim.fit)
```

```{r}
save.image("maize_mds.RData")
```

Note that due to the more or less random pick of marker covariates (cofactors), you might end up with different results from the function `cim()`. Therefore, it is important to save your R environment to use it in you final report. You can do so by using the function `save.image()`.

We will have better ways to verify whether there are other QTL on chromosomes 1 and 3 under the Multiple Interval Mapping (MIM) model approach.
