---
title: 'Week # 3 Class practice'
author: "Bashir
editor_options: 
  chunk_output_type: inline
---

---
,I. (2022)"
date: "24/03/2022"
output: html_document
---

We can use R to simulate what molecular markers might look like in a BC population using function sample.

```{r}
set.seed(123) # make sure results are reproduceable
mrk1 <- sample(x = c(0,1), size = 100, replace = TRUE, prob = c(0.5, 0.5))
table(mrk1)
```

Usually, we need to obtain marker information for hundreds or thousands of loci in order to have the genome covered by markers

```{r}
M <- 1000 # numebr of markers
N <- 100 # number of individuals
geno <- matrix(data = NA, nrow = N, ncol =M) # intialize an empty matrix to store the results
colnames(geno) <- paste0("Mrk", 1:M)
rownames(geno) <- paste0("Ind", 1:N)
geno[c(1:5), c(1:10)]
for  (m in 1:M) {geno[,m] <- sample(x =c(0,1), size = N, replace = TRUE, prob = c(0.5, 0.5))}
geno[c(1:5), c(1:10)] # show the 5 first rows and 10 first columns
```

```{r}
image(geno, ylab = "Markers", xlab = "Individuals", col = c("yellow", "blue"))
```

# Chi-Square tests (X^2^)

delete first row and first col

```{r}
data <- read.csv(file = "data/mouse.csv", header = TRUE)
data <- data[-1,-1]
head(data)
dim(data)
```

```{r}
M <- 14 # numebr of markers
N <- 103 # number of individuals

```

```{r}
exp <- c(0.5, 0.5) # the expected # the exepected proportion between categories in a BC and this proportion depend upon the type of marker e.g.  dominant , codominant or recessive markers
p.value <- rep(NA, M) # initializes an object to store the p-values
for (m in 1:M) {
  print(m)
  obs <- table(data[,m])
  p.value[m] <- chisq.test(x = obs, p= exp)$p.value
}
p.value

```

### Plotting

```{r}
plot(sort(p.value), ylim = c(0,1)) # plot of ordered p-values
abline(h = 0.05,  col = "red")
legend("topleft", legend=c("alpha=0.05"), col=c("red"), lty=1)
```

```{r}
plot(sort(p.value), ylim = c(0,1)) # plot of ordered p-values
abline(h = c(0.05, 0.05/M) ,  col = c("red", "blue"))
legend("topleft", legend=c("alpha=0.05", "alpha* = 0.05/M"), col=c("red", "blue"), lty=1)
```

## Multiple Test Correction

### Bonferroni correction

```{r}
table(p.value < 0.05/M)
plot(sort(p.value))
abline(h = c())
```

### Other correction methods

1.  **Per comparison error rate (PCER)**

2.  **Per-family error rate (PFER)**

3.  **Family-wise error rate (FWER)**

4.  **False discovery rate (FDR)**

5.  **Positive false discovery rate (*p*FDR)**
