---
title: "week_3_assignment"
author: "Bashir,I. (2022)"
date: "01/04/2022"
output: html_document
---

1.  In an R Markdown file, load the data files into R and compute the chi-square test using base R functions shown during the class "Chi-square tests, P-values and multiple testing correction (March 24, 2022)"

2.  Use Bonferroni correction to see whether marker loci follow the expected segregation ratio for each mouse and maize populations

3.  Submit the HTML file originated from your R Markdown script that you used to run the analyses above using the button "Add submission" below on this page

--------

# Data upload

```{r}
mouse <- read.csv(file = "data/mouse.csv", header = TRUE)
maize <- read.csv(file = "data/maize.csv", header = TRUE)
head(mouse)
head(maize)
```

# Data cleaning

```{r}
mouse <- mouse[-1,-1]
library(dplyr)
maize <- select(maize, -c('IND', 'GY'))
```

### Data structure

```{r}
dim(mouse)
dim(maize)
```

# Mouse Data Analysis
# P-values
```{r}
M <- 14 # number of markers
N <- 103 # number of individuals
```


```{r}
exp <- c(0.5, 0.5)
p.value <- rep(NA, M) # initializes an object to store the p-values
for (m in 1:M) {
  print(m)
  obs <- table(mouse[,m])
  p.value[m] <- chisq.test(x = obs, p= exp)$p.value}
p.value
```

```{r}
plot(sort(p.value), ylim = c(0,1)) # plot of ordered p-values
abline(h = 0.05,  col = "red")
legend("topleft", legend=c("alpha=0.05"), col=c("red"), lty=1)
```

## Multiple Test Correction

### Bonferroni correction

```{r}
plot(sort(p.value), ylim = c(0,1)) # plot of ordered p-values
abline(h = c(0.05, 0.05/M) ,  col = c("red", "blue"))
legend("topleft", legend=c("alpha=0.05", "alpha* = 0.05/M"), col=c("red", "blue"), lty=1)
```

# Maiza Data Analysis

```{r}
marker <- 12
number <- 171
exp1 <- c(0.25,0.5, 0.25) 
p.value1 <- rep(NA, marker) # initializes an object to store the p-values
for (m1 in 1:marker) {
  print(m1)
  obs <- table(maize[,m1])
  p.value1[m1] <- chisq.test(x = obs, p= exp1)$p.value
}
p.value1
```

```{r}
plot(sort(p.value1), ylim = c(0,1)) # plot of ordered p-values
abline(h = 0.05,  col = "red")
legend("topleft", legend=c("alpha=0.05"), col=c("red"), lty=1)
```

```{r}
plot(sort(p.value1), ylim = c(0,1)) # plot of ordered p-values
abline(h = c(0.05, 0.05/marker) ,  col = c("red", "blue"))
legend("topleft", legend=c("alpha=0.05", "alpha* = 0.05/marker"), col=c("red", "blue"), lty=1)
```
